NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-10-04 Manuscript/manuscriptPGPE.tex
Add todos
Update figures and text

\begin{table*}[htb]
  %\begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PE lipids.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systems}
  \begin{minipage}[t]{\textwidth}
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions  & NaCl (M) & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$   & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPE  & CHARMM36 \cite{??}           &0       & 144	& 5760  &0    & 310  & 500          & 400          & \cite{charmm36POPEfiles} \\
      POPE  & CHARMM36 \cite{??}           & 0      & 500       & 25000 & 0   &  310  & 500 & 100 & \cite{POPEcharmm} \\
      POPE  & CHARMM36 \cite{??}           & 0.11   & 500       & 25000 & 50  &  310  & 500 & 100 & \cite{POPEcharmm150mMNaCl} \\
      POPE  & CHARMM36ua \cite{??}         &0       & 336	& 15254 &0    & 310  & 2$\times$200 & 2$\times$100 & \cite{charmm36uaPOPEfiles}  \\
      \hline
      DPPE  & Slipids \cite{jambeck12b}    &0    & 288 	& 9386  &0    & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
      POPE  & Slipids \cite{jambeck12b,??} &0    & 336	& ?     &0    & 310  & 2$\times$200 &  2$\times$100 & \cite{slipidsPOPEfiles}  \\
      POPE  & Slipids \cite{??}            & 0    & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPEslipids} \\
      POPE  & Slipids \cite{??} \todoi{Ion parameters?}        & 0.11 & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPEslipids150mMNaCl} \\
      \hline
      DPPE  & GROMOS-CKP    \cite{??}      &0    & 128	& 3655  &0    & 342  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPdppe} \\
      POPE  & GROMOS-CKP    \cite{??}      &0    & 128	& 3552  &0    & 313  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPpope} \\
      POPE  & GROMOS-CKP    \cite{??}      &0    & 500	& 25000 &0    & 310  & 500 & 100 & \cite{gromosCKPpopeT310} \\
      DOPE  & GROMOS-CKP    \cite{??}      &0    & 128	& 4789  &0    & 271  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPdope} \\
      \hline
      POPE  & GROMOS 43A1-S3 \cite{??}     &0    & 128	& 3552     &0    & 313  & 2$\times$200 & 2$\times$100 & \cite{gromos43a1s3POPEfiles}  \\
      \hline
      POPE  & OPLS-UA vdW on H \cite{??}   &0    & 128	& 3328     &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{OPLSuaWvdWPOPEfiles} \\
      POPE  & OPLS-UA \cite{??}            &0    & 128	& 3328     &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{OPLSuaPOPEfiles} \\
      \hline
      POPE  & Berger-Vries \cite{??}       &0    & 128	& 3552  &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{bergerPOPEfiles}  \\
      POPE  & Berger-largeH \cite{??}      &0    & 128	& 3552  &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{berger2POPEfiles}  \\
      DOPE  & Berger-Vries \cite{??}       &0    & 128	& 4789  &0    & 271  & 2$\times$200 & 2$\times$100 & \cite{bergerDOPEfiles}  \\
      DOPE  & Berger-largeH \cite{??}      &0    & 128	& 4789  &0    & 271  & 2$\times$300 & 2$\times$100 & \cite{berger2DOPEfiles} \\ 
      \hline
      POPE             & LIPID17 \cite{??} & 0      & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17} \\
      POPE             & LIPID17 \cite{??} & 0.11   & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17150mMNaCl} \\
    \end{tabular}
  \end{minipage}
  %\end{sidewaystable*} 
\end{table*}

@@

@@

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}

The glycerol backbone order parameters in Slipid simulations are off from experiments
for both PE and PG lipids (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as already reported previously for PC lipids \cite{botan15} and PS lipids \cite{NMRlipidsIVps}.
The Slipid simulations were able to capture the essential differences between PC and PS lipid
headgroups \cite{NMRlipidsIVps}, but this is not the case for PE and PG lipids.
 \begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.pdf}
  \caption{\label{HGorderParametersPE}
    The headgroup and glycerol backbone order parameters of PE lipids
    from experiments (POPE and signs this work, DPPE from Ref.~\citenum{seelig76}
    and E.coliPE from Ref.~\citenum{gally81}) and simulations with different force fields.
  }
  \todo{This should be clarified as in NMRlipidsI and error bars should be added.
    Probably larger error bars for united atom models based on the report by Fuchs et al.
  }
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81}).
    and simulations with different force fields.
  }
\end{figure}

\todo{Berger results are not here yet, but we should mention the ring like structures
  pointed out by T. Piggot in the blog:
  The poor performance of headgroup order parameters in Berger model can be probably explained by ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}.
  These ring like structures are a widespread feature of typical Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
}

CHARMM36 simulations seems to capture all the essential differences between
headgroup order parameters of PC PE, PG (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as previously observed also for PS lipids \cite{NMRlipidsIVps}. Therefore,
the CHARMM36 simulations can be used to analyze the structural differences between
headgroups (Fig.~\ref{structures}).
\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/struct.png}
  \caption{\label{structures}
    Overlayed snapshots from CHARMM36 simulations of different lipids which give the best agreement with experiments.
  }
\end{figure}

\subsection{Lipid headgroup interactions in PC:PE and PC:PG mixtures}
According to the electrometer concept, the headgroup order parameters increase
with the addition of negatively charged PG lipids, but
%decrease when mixed with positively charged surfactants and
are not affected by the addition of zwitterionic PE lipids (Fig.~\ref{HGorderparametersPCvsPEPG}) \cite{seelig87, scherer87}. % or cholesterol.
%In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
%also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
%follow the eletrometer concept \cite{scherer87}.
%to mixtures of differently charged
%lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
%%\section{PC lipid headgroup response to different mixtures in experiments}
%As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
%headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
%i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
%surfactants. The changes with the addition of neutral lipids is significantly smaller.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of PC headgroup order parameters with increasing amount of PE (left) and PG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}
%The headgroup order parameters for PC lipids (POPC and DOPC)
%mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
%from different simulation model and experiments \cite{scherer87} with different
%mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
%follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
%parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
%remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is roughly observed also in simulations here, althouht the changes in PC headgroup with
PE are slightly overestimated and changes with PG lipid are slightly underestimated.
The underestimated increase of PC headgroup order parameters with the addition of
negaticely charged lipids was previously interepreted to indicate overestimated
counterion binding affinity, which overcompensates the effect of negative charge \cite{NMRlipidsIVps}.
Thus, the counterion binding to PG lipid bilayers (Fig. \ref{CIdensPG}) may be slightly
overestimated also in simulations here.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
  \caption{\label{CIdensPG}
    Counterion densities from simulations with PG lipids.
  }
\end{figure}

>>>>

  \caption{List of MD simulations with PE lipids.
  }\label{systems}
      lipid/counter-ions & force field for lipids / ions  & NaCl (M) & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$   & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      POPE  & CHARMM36 \cite{??}           &0       & 144	& 5760  &0    & 310  & 500          & 400          & \cite{charmm36POPEfiles} \\
      POPE  & CHARMM36 \cite{??}           & 0      & 500       & 25000 & 0   &  310  & 500 & 100 & \cite{POPEcharmm} \\
      POPE  & CHARMM36 \cite{??}           & 0.11   & 500       & 25000 & 50  &  310  & 500 & 100 & \cite{POPEcharmm150mMNaCl} \\
      POPE  & CHARMM36ua \cite{??}         &0       & 336	& 15254 &0    & 310  & 2$\times$200 & 2$\times$100 & \cite{charmm36uaPOPEfiles}  \\
      DPPE  & Slipids \cite{jambeck12b}    &0    & 288 	& 9386  &0    & 336  & 200 & 100 & \cite{slipidsDPPEfiles}  \\
      POPE  & Slipids \cite{jambeck12b,??} &0    & 336	& ?     &0    & 310  & 2$\times$200 &  2$\times$100 & \cite{slipidsPOPEfiles}  \\
      POPE  & Slipids \cite{??}            & 0    & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPEslipids} \\
      POPE  & Slipids \cite{??} \todoi{Ion parameters?}        & 0.11 & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPEslipids150mMNaCl} \\
      DPPE  & GROMOS-CKP    \cite{??}      &0    & 128	& 3655  &0    & 342  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPdppe} \\
      POPE  & GROMOS-CKP    \cite{??}      &0    & 128	& 3552  &0    & 313  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPpope} \\
      POPE  & GROMOS-CKP    \cite{??}      &0    & 500	& 25000 &0    & 310  & 500 & 100 & \cite{gromosCKPpopeT310} \\
      POPE  & GROMOS-CKP    \cite{??}      &0.11 & 500	& 25000 &50   & 310  & 500 & 100 & \cite{gromosCKPpopeT310150mMNaCl} \\
      DOPE  & GROMOS-CKP    \cite{??}      &0    & 128	& 4789  &0    & 271  & 2$\times$500 & 2$\times$400 & \cite{gromosCKPdope} \\
      POPE  & GROMOS 43A1-S3 \cite{??}     &0    & 128	& 3552     &0    & 313  & 2$\times$200 & 2$\times$100 & \cite{gromos43a1s3POPEfiles}  \\
      POPE  & OPLS-UA vdW on H \cite{??}   &0    & 128	& 3328     &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{OPLSuaWvdWPOPEfiles} \\
      POPE  & OPLS-UA \cite{??}            &0    & 128	& 3328     &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{OPLSuaPOPEfiles} \\
      POPE  & Berger-Vries \cite{??}       &0    & 128	& 3552  &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{bergerPOPEfiles}  \\
      POPE  & Berger-largeH \cite{??}      &0    & 128	& 3552  &0    & 303  & 2$\times$200 & 2$\times$100 & \cite{berger2POPEfiles}  \\
      DOPE  & Berger-Vries \cite{??}       &0    & 128	& 4789  &0    & 271  & 2$\times$200 & 2$\times$100 & \cite{bergerDOPEfiles}  \\
      DOPE  & Berger-largeH \cite{??}      &0    & 128	& 4789  &0    & 271  & 2$\times$300 & 2$\times$100 & \cite{berger2DOPEfiles} \\ 
      POPE             & LIPID17 \cite{??} & 0      & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17} \\
      POPE             & LIPID17 \cite{??} & 0.11   & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17150mMNaCl} \\

@@

  \todo{Data for POPC:POPG mixtures by listed by Antonio Peon is missing from this table}

@@

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
As reported previously for PC \cite{botan15} and PS \cite{NMRlipidsIVps} lipids,
the headgroup and glycerol backbone order parameters show wide variation between different force fields
for both PE and PG lipids (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
and none of the force fields reproduce all values within experimental error bars.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{NMRlipidsIVps}, but this is not the case for PE and PG lipids.
%\todo{Berger results are not here yet, but we should mention the ring like structures
%  pointed out by T. Piggot in the blog:
The poor performance of headgroup order parameters in Berger model can be probably explained by ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04},
which is a typical feature for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
%}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.pdf}
  \caption{\label{HGorderParametersPE}
    The headgroup and glycerol backbone order parameters of PE lipids
    from experiments (POPE and signs this work, DPPE from Ref.~\citenum{seelig76}
    and E.coliPE from Ref.~\citenum{gally81}) and simulations with different force fields.
  }
  \todo{This should be clarified as in NMRlipidsI and error bars should be added.
    Probably larger error bars for united atom models based on the report by Fuchs et al.
  }
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81}).
    and simulations with different force fields.
  }
\end{figure}
\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/struct.png}
  \caption{\label{structures}
    Overlayed snapshots from CHARMM36 simulations of different lipids which give the best agreement with experiments.
  }
\end{figure*}
Also in line with previous studies for PC \cite{botan15} and PS \cite{NMRlipidsIVps} lipids,
CHARMM36 simulations are in best agreement with experiments for headgroup and glycerol backbone
order parameters, and seem to capture all the essential differences between different headroups
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
%headgroup order parameters of PC PE, PG (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as previously observed also for PS lipids \cite{NMRlipidsIVps}.
In previous study, CHARMM36 predicted the more negative $\beta$-carbon order parameter and larger forking of
the $\alpha$-carbon is PS headgroup than in PC \cite{NMRlipidsIVps}.
In this work, the CHARMM36 simulations reproduce also the other essential differences in experimental
headgroup order parameters between different headgroups (Fig. \ref{HGorderParameters}) despite the inaccuracies
in individual segments:
The PE headgroup order parameters in CHARMM36 simulations are similar to the PC \cite{botan15}
and $\beta$-carbon order parameters is positive in PG headgroup, in contrast to the negative values observed in
other lipids. Therefore, we use CHARMM36 simulations to analyze the structural differences between
headgroups (Fig.~\ref{structures}). While rotation around N-C$_\alpha$-C$_\beta$-O$_\alpha$ dihedral was
significantly different in PS and PC headgroups in previous work \cite{NMRlipidsIVps}, essential differences
between PC, PE and PG are not observed here (Fig.~\ref{structures}) \todo{We need also the dihedral distributions to finish this discussion.}. \\
\todo{Why is the $\beta$-carbon order parameter of PG different to PC and PE then?}

\subsection{Lipid headgroup interactions in PC:PE and PC:PG mixtures}
According to the electrometer concept, the headgroup order parameters increase
with the addition of negatively charged PG lipids, but
%decrease when mixed with positively charged surfactants and
are not affected by the addition of zwitterionic PE lipids (Fig.~\ref{HGorderparametersPCvsPEPG}) \cite{seelig87, scherer87}. % or cholesterol.
%In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
%also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
%follow the eletrometer concept \cite{scherer87}.
%to mixtures of differently charged
%lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
%%\section{PC lipid headgroup response to different mixtures in experiments}
%As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
%headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
%i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
%surfactants. The changes with the addition of neutral lipids is significantly smaller.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of PC headgroup order parameters with increasing amount of PE (left) and PG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Data for CHARMM, Lipid17, Slipids, and also GROMOS also without ions is now available.
    This data should be updated into this plot}
  %\todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}
%The headgroup order parameters for PC lipids (POPC and DOPC)
%mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
%from different simulation model and experiments \cite{scherer87} with different
%mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
%follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
%parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
%remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is roughly observed also in simulations here, althouht the changes in PC headgroup with
PE are slightly overestimated and changes with PG lipid are slightly underestimated.
The underestimated increase of PC headgroup order parameters with the addition of
negaticely charged lipids was previously interepreted to indicate overestimated
counterion binding affinity, which overcompensates the effect of negative charge \cite{NMRlipidsIVps}.
Thus, the counterion binding to PG lipid bilayers (Fig. \ref{CIdensPG}) may be slightly
overestimated also in simulations here.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
  \caption{\label{CIdensPG}
    Counterion densities and area per lipids from simulations with PG lipids.
    Experimental area for POPG at 303 K is 66.1 {\AA}$^2$ and 67 {\AA}$^2$ for DPPC at 323 K \cite{pan12b}.
  }
  \todo{Data for GROMOS is also available without ions, but I could not analyze it, probably because tpr file was not correct.}
\end{figure}
